---
title: "Filtering FinBIF records"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering FinBIF records}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, eval=TRUE, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache.path = "cache/"
)
```
When getting records from FinBIF there are many options for filtering the data
before it is downloaded saving bandwidth and local post-processing time. For the
full list of filtering options see `?filters`.

## Loading finbif
```{r setup}
library(finbif)
```

## Informal taxonomic groups
You can filter occurrence records based on informal taxonomic groups such as
`Birds` or `Mammals`.
```{r informal-groups, cache = TRUE}
finbif_occurrence(filter = list(informal_group = c("Birds", "Mammals")))
```

See `finbif_informal_groups()` for the full list of groups you can filter by. 
You can use the same function to see the subgroups that make up a higher
level informal group:
```{r informal-subgroups}
finbif_informal_groups("macrofungi")
```

## Administrative status
Many records in the FinBIF database include taxa that have one or another
administrative statuses. See `finbif_admin_status()` for a list of
administrative statuses and short-codes.
```{r admin-status, cache = TRUE}
# Search for birds on the EU invasive species list
finbif_occurrence(
  filter = list(informal_group = "Birds", administrative_status = "EU_INVSV")
)
```

## IUCN red list
Filtering can be done by [IUCN red list](https://www.iucnredlist.org) 
category. See `finbif_red_list()` for the IUCN red list categories and their 
short-codes.
```{r red-list-status, cache = TRUE}
# Search for near threatened mammals
finbif_occurrence(
  filter = list(informal_group = "Mammals", red_list_status = "NT")
)
```

## Habitat type
Many taxa are associated with one or more primary or secondary habitat types
(e.g., forest) or subtypes (e.g., herb-rich alpine birch forests). Use
`finbif_habitat_types()` to see the habitat types in FinBIF. You can filter
occurrence records based on primary (or primary/secondary) habitat type or
subtype codes. Note that filtering based on habitat is on taxa not on the
location (i.e., filtering records with `primary_habitat = "M"` will only return
records of taxa considered to primarily inhabit forests, yet the locations of
those records may encompass habitats other than forests).
```{r habitat-type, cache = TRUE}
head(finbif_habitat_types())

# Search records of taxa for which forests are their primary or secondary
# habitat type
finbif_occurrence(filter = c(primary_secondary_habitat = "M"))
```

You may further refine habitat based searching using a specific habitat type
qualifier such as "sun-exposed" or "shady". Use `finbif_habitat_qualifiers()`
to see the qualifiers available. To specify qualifiers use a named list of
character vectors where the names are habitat types or subtypes and the elements
of the character vectors are the qualifier codes.
```{r habitat-type-specific, cache = TRUE}
finbif_habitat_qualifiers()[4:6, ]

# Search records of taxa for which forests with sun-exposure and broadleaved
# deciduous trees are their primary habitat type
finbif_occurrence(filter = list(primary_habitat = list(M = c("PA", "J"))))
```
