---
title: "Occurrence records from FinBIF"
author: "William K. Morris"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2. Occurrence records from FinBIF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, eval = TRUE, echo = FALSE, cache = FALSE}
library(details)
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "cache/",
  fig.path = "../man/figures/",
  details.summary = "Click to show/hide output.",
  details.tooltip = NULL,
  details.comment = "#>"
)

library(finbif)
options(finbif_tz = "Europe/Helsinki", width = 100)
```

The core purpose of `{finbif}` is accessing occurrence data stored in the FinBIF
database. Occurrence data can be retrieved from FinBIF with the function
`finbif_occurrence()`. Without any arguments specified `finbif_occurrence()`
will retrieve the latest 10 occurrence records from FinBIF.
```{details occurrence_vanilla}
finbif_occurrence()
```
The print method for the resulting `finbif_occ` object will display the number
of records downloaded, the total number of records available, a data summary
including up to 10 rows of some core record variables (when available), the
number of remaining records and variables, as well as the names of additional
variables.

## Darwin Core Variables
You can switch from the default variable names to [Darwin Core
](http://rs.tdwg.org/dwc/) style names by setting `dwc = TRUE`.
```{r dwc}
colnames(finbif_occurrence(dwc = TRUE))
```
The functions `to_dwc()` and `to_native()` can be used to translate variable
names to and from Darwin Core style and `{finbif}`'s native variable names
style.

## Choosing taxa
You can limit the records to certain taxa by specifying them as an argument.
```{details cygnus}
finbif_occurrence("Cygnus cygnus")
```

Multiple taxa can be requested at once.
```{details cygnus2}
finbif_occurrence("Cygnus cygnus", "Cygnus olor")
```

You can also chose higher taxonomic groups and use common names (in English,
Finnish and Swedish).
```{r birds}
birds  <- finbif_occurrence("Birds")
linnut <- finbif_occurrence("Linnut")
faglar <- finbif_occurrence("Fåglar")

sapply(list(birds, linnut, faglar), nrow)
```

## Request size
You can increase the number of records returned by using the `n` argument.
```{details occurrences1001}
finbif_occurrence(n = 1001)
```

You can see how many records are available for a given request, without
retrieving any records, by setting `count_only = TRUE`.
```{r count-only}
finbif_occurrence(count_only = TRUE)
```

## Checking taxa
When you request occurrence records for specific taxa, by default, the taxon
names are first checked against the FinBIF database. If any of the requested
taxa are not found in the database you will receive a warning but the data will
still be retrieved for the remaining taxa.
```{details moomin}
finbif_occurrence("Vulpes vulpes", "Moomin")
```

You can turn off taxon name pre-checking by setting the value of the
`check_taxa` argument to `FALSE`.
```{details moomin-check-false}
finbif_occurrence("Vulpes vulpes", "Moomin", check_taxa = FALSE)
```

By setting the argument, `on_check_fail` to `"error"` (the default is `"warn"`),
you can elevate the warnings to errors and the request will fail if any of the
taxa are not found in the FinBIF database.
```{r moomin-check-error, error = TRUE}
finbif_occurrence("Vulpes vulpes", "Moomin", on_check_fail = "error")
```
This can be a useful strategy if you are using `{finbif}` non-interactively
(in a script), and you do not want to proceed if any of your taxon names are
wrong or misspelled.

## Aggregating records
You can request records in aggregate using the `aggregate` argument to
`finbif_occurrence`. Aggregated requests will return counts for the combination
of the `variables` you specify with the `select` argument. You can request
counts of `"records"`, `"species"` or `"taxa"` by using the corresponding string
as the value for the `aggregate` argument. Aggregating by `"species"` will count
the number of unique species identifiers for a set of records grouped by the
combination of selected variables. Note that this count will not include records
of taxa that do not have species identifiers, including records of higher taxa
(e.g., genus only records), records of the non-species children of aggregate or
complex taxa, and hybrid taxa. Therefore, in some contexts the results returned
will be an underestimate of species richness. Likewise, aggregating by `"taxa"`,
which returns a count the number of unique taxon identifiers, could represent an
overestimate of the number of taxa as records of higher taxa will contribute to
the count while their true identify may be a duplicate of other records.

To illustrate, you can count the number of moths and butterflies by municipality
with the following:
```{details aggregate}
finbif_occurrence(
  "Lepidoptera", select = "finnish_municipality", aggregate = "species"
)
```

## Time & duration
The default behaviour of `finbif_occurrence` is to consolidate date and time
data for occurrence recording events into a `date_time` variable. This can be
turned off (which can speed up data processing time) by deselecting the
`date_time` variable.
```{details date-time-false}
finbif_occurrence(select = "-date_time")
```

### Timezone
#### Timezone input
The FinBIF database doesn't currently store timezone information, so `{finbif}`
makes assumptions about the appropriate timezone based on the time and location
of the occurrence recording events to calculate `date_time` and `duration`. By
default, a fast heuristic is used to determine the timezones. If you require
greater accuracy (e.g., you are using data on the Finnish/Swedish border and
daytime/nighttime hours are important), you can switch to more accurate, though
slower, timezone calculation method.
```{details date-time-accurate}
finbif_occurrence(date_time_method = "accurate")
```

#### Timezone output
The timezone of the calculated `date_time` variable is determined by the
timezone of your operating system.
```{r sys-tzone, eval = FALSE, cache = FALSE}
Sys.timezone()
```

You can override this by setting the `tzone` argument to a different value.
```{details tzone}
finbif_occurrence(tzone = "Etc/UTC")
```

Or set the global timezone option to set the timezone for the current session.
```{r tzone-global, eval = FALSE, cache = FALSE}
options(finbif_tz = "Etc/UTC")
```
This may be advisable for reproducibility or when working with multiple systems.
