---
title: "Filtering occurrence records"
author: "William K. Morris"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{5. Filtering occurrence records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, eval = TRUE, echo = FALSE, cache = FALSE}
library(details)
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "cache/",
  fig.path = "../man/figures/",
  details.summary = "Click to show/hide output.",
  details.tooltip = NULL,
  details.comment = "#>"
)

library(finbif)
options(finbif_tz = "Europe/Helsinki", width = 100)
```
When getting records from FinBIF there are many options for filtering the data
before it is downloaded, saving bandwidth and local post-processing time. For
the full list of filtering options see `?filters`.

## Location
Records can be filtered by the name of a location or by a set of coordinates.
```{r location-filter, fig.cap = ""}
filter1 <- c(country = "Finland")
filter2 <- list(coordinates = list(c(60, 68), c(20, 30), "wgs84"))
par(mfcol = 1:2)
plot(finbif_occurrence(filter = filter1, n = 1000), main = "Name")
plot(finbif_occurrence(filter = filter2, n = 1000), main = "Coordinates")
```

See `?filters` section "Location" for more details

## Time
The event or import date of records can be used to filter occurrence data from
FinBIF. The date filters can be a single year, month or date,
```{details date-filter}
finbif_occurrence(filter = list(date_range_ym = c("2019-12")))
```
, or for record events, a range as a character vector or an
[Interval](https://lubridate.tidyverse.org/reference/Interval-class.html)
object.
```{details date-range-filter}
finbif_occurrence(
  filter = list(date_range_ymd = c("2019-06-01", "2019-12-31"))
)
```

Records for a specific season or time-span across all years can also be
requested.
```{details season-filter}
finbif_occurrence(
  filter = list(
    date_range_md = c(begin = "12-21", end = "12-31"),
    date_range_md = c(begin = "01-01", end = "02-20")
  )
)
```

## Data Quality
You can filter occurrence records by indicators of data quality. See `?filters`
section "Quality" for details.
```{r quality-filter}
strict <- c(
  collection_quality = "professional", coordinates_uncertainty_max = 1,
  record_quality = "expert_verified"
)
permissive <- list(
  quality_issues = "both",
  record_reliability = c("reliable", "unassessed", "unreliable"),
  record_quality = c(
    "expert_verified", "community_verified", "unassessed", "uncertain",
    "erroneous"
  )
)
c(
  strict     = finbif_occurrence(filter = strict,     count_only = TRUE),
  permissive = finbif_occurrence(filter = permissive, count_only = TRUE)
)
```

## Collection
The FinBIF database consists of a number of constituent collections. You can
filter by collection with either the `collection` or `not_collection` filters.
Use `finbif_collections()` to see metadata on the FinBIF collections.
```{r collection-filter}
finbif_occurrence(
  filter = c(collection = "iNaturalist Suomi Finland"), count_only = TRUE
)
finbif_occurrence(
  filter = c(collection = "Notebook, general observations"), count_only = TRUE
)
```

## Informal taxonomic groups
You can filter occurrence records based on informal taxonomic groups such as
`Birds` or `Mammals`.
```{details informal-groups-filter}
finbif_occurrence(filter = list(informal_groups = c("Birds", "Mammals")))
```

See `finbif_informal_groups()` for the full list of groups you can filter by.
You can use the same function to see the subgroups that make up a higher
level informal group:
```{r informal-subgroups}
finbif_informal_groups("macrofungi")
```

## Regulatory 
Many records in the FinBIF database include taxa that have one or another
regulatory statuses. See `finbif_metadata("regulatory_status")` for a list of
regulatory statuses and short-codes.
```{details regulatory-statuses}
# Search for birds on the EU invasive species list
finbif_occurrence(
  filter = list(informal_groups = "Birds", regulatory_statuses = "EU_INVSV")
)
```

## IUCN red list
Filtering can be done by [IUCN red list](https://punainenkirja.laji.fi/)
category. See `finbif_metadata("red_list")` for the IUCN red list categories and
their short-codes.
```{details red-list-status}
# Search for near threatened mammals
finbif_occurrence(
  filter = list(informal_groups = "Mammals", red_list_status = "NT")
)
```

## Habitat type
Many taxa are associated with one or more primary or secondary habitat types
(e.g., forest) or subtypes (e.g., herb-rich alpine birch forests). Use
`finbif_metadata("habitat_type")` to see the habitat types in FinBIF. You can
filter occurrence records based on primary (or primary/secondary) habitat type
or subtype codes. Note that filtering based on habitat is on taxa not on the
location (i.e., filtering records with `primary_habitat = "M"` will only return
records of taxa considered to primarily inhabit forests, yet the locations of
those records may encompass habitats other than forests).
```{r habitat-type-head}
head(finbif_metadata("habitat_type"))
```
```{details habitat-type-filter}
# Search records of taxa for which forests are their primary or secondary
# habitat type
finbif_occurrence(filter = c(primary_secondary_habitat = "M"))
```

You may further refine habitat based searching using a specific habitat type
qualifier such as "sun-exposed" or "shady". Use
`finbif_metadata("habitat_qualifier")` to see the qualifiers available. To
specify qualifiers use a named list of character vectors where the names are
habitat types or subtypes and the elements of the character vectors are the
qualifier codes.
```{r habitat-type-specific-meta}
finbif_metadata("habitat_qualifier")[4:6, ]
```
```{details habitat-type-specific-filter}
# Search records of taxa for which forests with sun-exposure and broadleaved
# deciduous trees are their primary habitat type
finbif_occurrence(filter = list(primary_habitat = list(M = c("PA", "J"))))
```

## Status of taxa in Finland
You can restrict the occurrence records by the status of the taxa in Finland.
For example you can request records for only rare species.
```{details occurrence-status-filter}
finbif_occurrence(filter = c(finnish_occurrence_status = "rare"))
```
Or, by using the negation of occurrence status, you can request records of birds
excluding those considered vagrants.
```{details occurrence-status-neg-filter}
finbif_occurrence(
  filter = list(
    informal_groups               = "birds",
    finnish_occurrence_status_neg = sprintf("vagrant_%sregular", c("", "ir"))
  )
)
```

See `finbif_metadata("finnish_occurrence_status")` for a full list of statuses
and their descriptions.
