---
title: "Filtering FinBIF records"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering FinBIF records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When getting records from FinBIF there are many options for filtering the data
before it is downloaded saving bandwidth and local post-processing time. For the
full list of filtering options see `?filters`.

## Loading finbif
```{r setup}
library(finbif)
```

## Informal taxonomic groups
You can filter occurrence records based on informal taxonomic groups such as
`Birds` or `Mammals`.
```{r informal-groups}
finbif_occurrence(filters = list(informal_group = c("Birds", "Mammals")))
```

See `finbif_informal_groups()` for the full list of groups you can filter by. 
You can use the same function to see the subgroups that make up a higher
level informal group:
```{r informal-subgroups}
finbif_informal_groups("macrofungi")
```

## Administrative status
Many records in the FinBIF database include taxa that have one or another
administrative statuses. See `finbif_admin_status()` for a list of
administrative statuses and short-codes.
```{r admin-status}
# Search for birds on the EU invasive species list
finbif_occurrence(
  filters = list(informal_group = "Birds", administrative_status = "EU_INVSV")
)
```

## IUCN red list
Filtering can be done by [IUCN red list](https://www.iucnredlist.org) 
category. See `finbif_red_list()` for the IUCN red list categories and their 
short-codes.
```{r red-list-status}
# Search for near threatened mammals
finbif_occurrence(
  filters = list(informal_group = "Mammals", red_list_status = "NT")
)
```
